---
title: "Summarising data with `dplyr` and pipes `%>%`"
subtitle: "Practical 11"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---



```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
library(dplyr)
library(palmerpenguins)
penguins2 <- penguins %>% select(species, island, sex, year)
style_duo_accent(
  primary_color = "#003b49",
  secondary_color = "#1d4289",
  header_font_google = google_font("Cabin"),
  text_font_google = google_font("Noto Sans", "300", "300i"),
  code_font_google = google_font("Fira Mono"),
  colors = c(
    red = "#d3003f",
    purple = "#3e2f5b",
    orange = "#ff8811",
    green = "green",
    white = "#FFFFFF",
    blue = "blue"
  )
)

xaringanthemer::style_extra_css(
  list(
    ".title-slide" = list(
      "background-image" =
        paste0(
          "url(https://upload.wikimedia.org/wikipedia/commons/thumb/3/34/",
          "University_of_Sussex_Logo.svg/",
          "480px-University_of_Sussex_Logo.svg.png)"
        ),
      "background-position" = "95% 95%",
      "background-size" = "180px",
      "border" = "10px solid #013035",
      "background-color" = "#FFFFFF"
    ),
    ".title-slide h1" = list(
      "padding-top" = "0px",
      "font-size" = "60px",
      "text-align" = "left",
      "padding-bottom" = "18px",
      "margin-bottom" = "18px",
      "margin-top" = "0px",
      "color" = "#003b49"
    ),
    ".title-slide h2" = list(
      "font-size" = "40px",
      "text-align" = "left",
      "padding-top" = "10px",
      "margin-top" = "0px",
      "color" = "#003b49"
    ),
    ".title-slide h3" = list(
      "font-size" = "30px",
      "color" = "#26272A",
      "text-align" = "left",
      "text-shadow" = "none",
      "padding" = "0px",
      "margin" = "0px",
      "line-height" = "1"
    ),
    ".big" = list("font-size" = "40px"),
    "pre" = list(
      "background" = "#f8f8f8",
      "padding" = "10px",
      "border" = "1px",
      "border-color" = "#666",
      "border-style" = "solid",
      "border-radius" = "2px"
    ),
    ".footnote" = list(
      "font-size" = ".6em",
      "border-top" = "1px solid #007935"
    )
  )
)


xaringanExtra::use_editable(expires = 1)
```
```{css echo=FALSE}

  kbd {
      padding: 2px 4px;
      font-size: 90%;
      color: rgb(var(--font-col));
          background-color: #efefef;
      border-radius: 3px;
      box-shadow: none;
      border: solid 1px;
  }
```

```{r, echo=FALSE}
library(metathis)
week <- "10"
description <- paste0(
  c(
  "In this practical we'll learn how to group data, ",
  "summarise data, and use pipes `%>%` to chain operations together")
)
content_type <- "intro_slides"

meta() %>%
  meta_description(description) %>%
  meta_tag("week" = week) %>%
  meta_tag("content_type" = content_type)
```


## Plan for today

- Questions about last week's practical 

- Attendance pin

- More on manipulating and transforming data with `dplyr`

  - the `dplyr::summarise()` function

  - the `dplyr::group_by()` function

- Chaining together operations with `%>%`


---


# Attendance 

<br>

.middle[
.center[

## Attendance pin.can-edit[: ]



<br>
<br>
<br>
<br>
<br>
<br>
<br>

**Go to www.sussex.ac.uk/mobile OR use the SussexMobile app to input the pin**

]]


---

## Recap of `dplyr::mutate()`

Last we say how we could use the `dplyr::mutate()` function to add a new column
that held an average of another column. 


Let's say we wanted to create a new column called `average_price`, which
contains the average of the values in the price column. 

```{r, echo=FALSE, include=FALSE}

fruit_prices <- tibble::tibble(item = c("Apples", "Bananas", "Oranges"), price = c(1.00, 2.00, 2.40))
```

.pull-left[
Our original tibble with fruit prices

```{r, include=TRUE, echo=FALSE}
fruit_prices
```
]


.pull-right[
Our tibble with the average price column

```{r, include=TRUE, echo=FALSE}
dplyr::mutate(.data = fruit_prices, average_price = mean(price))
```
]

We'd run this command:

```r
dplyr::mutate(.data = fruit_prices, average_price = mean(price))
```

---

## The `dplyr::summarise()` function

But let's say that we want to collapse our table down to make a summary table.

That is, we don't want the whole table. We just one row with the average

.pull-left[
Our tibble with the average price column

```{r, include=TRUE, echo=FALSE}
dplyr::mutate(.data = fruit_prices, average_price = mean(price))
```
]


.pull-right[
A summary with only average price

```{r, include=TRUE, echo=FALSE}
dplyr::summarise(.data = fruit_prices, average_price = mean(price))
```
]

To do this, we'd use the the `dplyr::summarise()` function

```r
dplyr::summarise(.data = fruit_prices, average_price = mean(price))
```

Notice the identical syntax for `dplyr::mutate()` and `dplyr::summarise()`

---

## Adding multiple of summarises 

In the previous example we just added a single summary column, but we can add
multiple ones

.pull-left[
Our orginal **tibble** with `price`

```{r, include=TRUE, echo=FALSE}
fruit_prices
```
]


.pull-right[
Two summarises for `mean` and `sd`

```{r, include=TRUE, echo=FALSE}
dplyr::summarise(
  .data = fruit_prices, mean_price = mean(price),
  sd_price = sd(price)
)
```
]

To do this, we'd use the the `dplyr::summarise()` function

```r
dplyr::summarise(.data = fruit_prices, mean_price = mean(price), 
                                       sd_price = sd(price))

```

If you know how to use `dplyr::mutate()` then `dplyr::summarise()` is easy to
learn!

---

## The `dplyr::group_by()` function

The `dplyr::summarise()` function is not that useful **by itself**. It's real
power comes when it's combined with `dplyr::group_by()`, which allows us to
**group together rows** according to a **grouping variable**

```{r, include=FALSE}
sub_ids <- rep(paste0("sub", sprintf("%.2d", seq(1, 10, 1))), 2)
cond <- c(rep("inc", 10), rep("con", 10))
rt <- round(c(rnorm(10, 900, 50), rnorm(10, 800, 50)), 2)
stroop <- tibble::tibble(id = sub_ids, condition = cond, rt = rt) %>% arrange(id)
```


.pull-left[
A tibble that hasn't been grouped
```{r, include=TRUE, echo=FALSE}
print(stroop, n = 4)
```
]

.pull-right[
...and now grouped by **condition**
```{r, include=TRUE, echo=FALSE}
print(group_by(stroop, condition), n = 3)
```
]

To **group** the tibble **by condition** we'd just run
```r
dplyr::group_by(.data = stroop, condition)
```
---
class: smoller

### Working with grouped data


Once we have grouped data, then we can calculate summaries **by group**. 

.pull-left[
A tibble grouped by condition 
```{r, include=TRUE, echo=FALSE}
print(group_by(stroop, condition), n = 4)
```
]

.pull-right[
A summary of a grouped tibble
```{r, include=TRUE, echo=FALSE}
print(group_by(stroop, condition) %>% summarise(mean_rt = mean(rt), .groups = "drop"), n = 4)
```
]

To do this would take **two steps**...

```r
# First we group it!
grouped_stroop <- dplyr::group_by(.data = stroop, condition)

# Then we summarise it!
dplyr::summarise(.data = grouped_stroop, mean_rt = mean(rt))

```

---

### Ungrouping data with `dplyr::ungroup()`

One this to be aware of when you're working with **grouped** data is that your 
data will stay grouped

Once you're finished working with the data in **grouped** form, you should
**explicitly** ungroup it, otherwise you may sometimes get unexpected results!

To **ungroup** a **tibble** just use `dplyr::ungroup()`

```r
ungrouped_stroop <- dplyr::ungroup(.data = grouped_stroop)
```

---

### Useful functions for summarises

There are a few functions that are particularly useful for using with
`dplyr::summarise()` (and  `dplyr::group_by()`)

- `mean(..., na.rm = TRUE)` for working out the mean of a **column's values**

- `sd(..., na.rm = TRUE)` for working out standard deviation of a **column's
  values** 

- `n()` for counting the **numbers of rows** (in a group if the data is
  grouped)

For the `mean()` and `sd()` we need to specify the **column name** we want to
work out the mean or sd of

If the column contains **missing values** then we can add `na.rm = TRUE` to
ignore those missing values

For the `n()` function you **don't** specify a column name, because *all the
columns have the same number of rows* so any input would be redundant 

---

## Using `%>%` to chain together steps

The `%>%` is called the **pipe operator**. We use it for **piping data
around!**

The `%>%` doesn't come built in to `R` but it is part of a package called
`magrittr`<sup>1</sup>

To be able to access the `%>%` operator you first need to install `magrittr`

```r
install.packages("magrittr")
```

And then load it 

```r
library("magrittr")
```

Everyone should already have it installed, so you should only need to load it!

.footnote[<sup>1</sup>It's named after René Magritte, the Belgian surrealist
painter who famous for the artwork [The Treachery of Images
](https://en.wikipedia.org/wiki/The_Treachery_of_Images), which is picture of a
pipe with the text "Ceci n'est pas une pipe" (translation: "This is not a
pipe")]

---

## Using `%>%` to chain together steps

*The two step version*

<pre><code>
<span style="background-color:lightblue; font-weight:bold">grouped_stroop</span> <- dplyr::group_by(.data = stroop, condition)

dplyr::summarise(.data = <span style="background-color:lightblue; font-weight:bold">grouped_stroop</span>, mean_rt = mean(rt))
</code></pre>


The **output** from *step 1* serves as the **input** for *step 2* 

*The one step version*

<pre><code>
dplyr::group_by(.data = stroop, condition) <span style="background-color:lightgreen; font-weight:bold">%>%</span>
      dplyr::summarise(.data = <span style="background-color:lightblue; font-weight:bold">.</span>, mean_rt = mean(rt))
</code></pre>

We don't **assign the output** from *step 1* to an object. Instead, we add a
`%>%` and then start writing out *step 2*, and we use `.` to indicate where the
**output** from *step 1* should go

The two steps now function *like a single command*

---


## Using `%>%` to chain together steps

You can chain together **any number of steps** and use any functions you want!

You can also move the first input outside the first function and pipe that in too!

Let's say you start off with something like this:

```{r, echo=FALSE}
penguins %>% select(species, island, body_mass_g, bill_length_mm, sex, year)
```

---
class: smoller

```{css, echo=FALSE}
.smoller .remark-code {
  font-size: 80%;
}
```

And you want something like this:

```{r, echo=FALSE}
penguins %>%
  dplyr::filter(.data = ., sex == "female") %>%
  dplyr::group_by(.data = ., species) %>%
  summarise(
    .data = .,
    avg_mass = mean(body_mass_g),
    sd_mass = sd(body_mass_g),
    n = n(), .groups = "drop"
  ) %>%
  mutate(.data = ., se = sd_mass / sqrt(n))
```

The **mean**, the **standard deviation**, the **number of penguins**, and the
**standard error of the mean** for the **females** of each of each **species**

```r
penguins %>% 
      dplyr::filter(.data = ., sex == 'female') %>%  
      dplyr::group_by(.data = ., species) %>%  
      summarise(.data = .,  
            avg_mass = mean(body_mass_g),
            sd_mass = sd(body_mass_g), 
            n = n()) %>%  
      mutate(.data = ., se = sd_mass/sqrt(n)) 
```

This seems **very complicated now** but once you get used to *breaking things
down into steps joined with `%>%`* it will start to look easier

---
class: smoller

And you want something like this:

```{r, echo=FALSE}
penguins %>%
  dplyr::filter(.data = ., sex == "female") %>%
  dplyr::group_by(.data = ., species) %>%
  summarise(
    .data = .,
    avg_mass = mean(body_mass_g),
    sd_mass = sd(body_mass_g),
    n = n(), .groups = "drop"
  ) %>%
  mutate(.data = ., se = sd_mass / sqrt(n))
```

The **mean**, the **standard deviation**, the **number of penguins**, and the
**standard error of the mean** for the **females** of each of each **species**

```r
penguins %>% 
      dplyr::filter(.data = ., sex == 'female') %>% # filter for females 
      dplyr::group_by(.data = ., species) %>%  # group by species
      summarise(.data = .,  # work out summarises
            avg_mass = mean(body_mass_g), # the mean
            sd_mass = sd(body_mass_g), # the sd
            n = n()) %>%  # the count
      mutate(.data = ., se = sd_mass/sqrt(n)) # calculate the standard error
```

This seems **very complicated now** but once you get used to *breaking things
down into steps joined with `%>%`* it will start to look easier

---

# Tips for using  `%>%`

There are two main tips for using `%>%`

- Build up pipelines *step by step* so you can keep track of what each
  transformation is doing 

- You only ever need to make them as complex as you want to!
