---
title: "Introduction to R Markdown"
subtitle: "Tutorial 03"
author: "Psychology as a Science"
output:
  rmarkdown::html_document:
    css: ["./css/paas_learnr.css", "./css/katex.min.css"] 
    includes: 
      js: "./js/setup.js"
      after_body: "./header.html"
    self_contained: false
---
 


```{r, include=FALSE}
library(metathis)
library(niceQuiz)

knitr::opts_chunk$set(echo = FALSE)
meta() %>%
  meta_description("Week 3 Tutorial") %>%
  meta_tag("content_type" = "tutorial",
           "week" = "03")
```

```{css}
.verbatim > * {word-break:break-word}
````

## Intro

Welcome to the third PAAS tutorial, in which we will talk about text editors.
We will contrast the word processing software that you are probably used to
with markdown, a slightly geekier but way of writing documents. Finally, we
will discover RStudio's own markdown editor and learn how to write documents
using it.

\ 

Onwards!

\ 

## Text files

Text files are simply computer files that contain text. To create or modify
them, we need programs called text editors. Remember how we said that the .docx
files require Microsoft Word (or similar) to be installed on the device in
order to open them. Well, the .docx file is a text file and MS Word is the text
editor program. There are two basic kinds of text, plain text and rich text.


### Plain text

Plain text is basically what it says on the tin: just text, without formatting.
There is no **bold text**, no *italics*, no ^superscripts^, no
<big>variable</big> <small>font sizes</small>, just plain text.

A typical plain text file has the .txt file extension. This extension lets the
computer know that the contents of the file are just text and should not
necessarily be interpreted as anything else. However, .txt are not the only
plain text files. Later on, when we are writing longer blocks of `R` code, we
will be saving them into files with a .R extension. These are still plain text
files but, when opened in the source pane of RStudio, the program will decorate
the text with different colours for different bits of the code to make the code
easier for us to read.


### Rich text

Rich text is text with all sorts of formatting. What you are reading now is
rich text. You can see that there are headings, buttons, fonts of different
colours and sizes, and so on...

Rich text formatting is, however, just a visual representation that the program
creates for us and shows it on the screen. Behind the scenes the program we are
using is reading a plain text file written using one of several formatting
systems. Every one of these systems has its own way to indicate, using plain
text symbols and tags, how the displayed text should be formatted. For example,
.docx files use a system called [Extensible Markup Language, or
XML](https://en.wikipedia.org/wiki/XML), while Google Docs use [Hypertext
Markup Language, or HTML](https://en.wikipedia.org/wiki/HTML).

HTML is the system in which modern websites are written and all web browsers
know how to turn HTML plain text into formatted rich text. These tutorials are
also written in HTML and so can be opened in your web browser. To illustrate
what we mean by indicating formatting using plain text, let's see what the HTML
plain text behind the first paragraph of the "Plain text" section above looks
like:

<div class="verbatim">
```
<p>Plain text is basically what it says on the tin: just text, without formatting. There is no <strong>bold text</strong>, no <em>italics</em>, no <sup>superscirpts</sup>, no <big>variable</big> <small>font sizes</small>.</p>
```
</div>

The browser reading this text knows that this is a new paragraph
(`<p>...</p>`), the "bold text" bit should be bold because it's inside the
`<strong>...</strong>` tag, *etc*.

Again, there's no magic here, nothing mysterious is happening, just a set of
instructions in plain text that the program (the browser, in this case) knows
how to follow.

\ 

Now that we understand how text files work, let's see how we can create them.

\ 

## Text editors

You are probably familiar with MS Word and Google Docs and maybe you use them
to write documents. These two are examples of text editors, programs that cn
create and modify text files. Because all text formatting and styling in these
editors is done in real time using buttons, menus, or keyboard shortcuts so
that you are always seeing the actual formatted version of the final document,
these kinds of editors are called "WYSIWYG" editors (What You See Is What You
Get).

There is also a different kind of text editors that precedes WYSIWYG - plain
text editors. As their name suggests, they only deal with plain text. Using
this approach to creating and editing documents involves two steps:

1. You write the plain text file, indicating the structure of the text using
   symbols or tags.
2. Once the plain text file is written, the editor converts it to a formatted
   file, such as .docx, .html, or .pdf.

There are several "markup languages" that can be used to write the plain text
but the one we will be using is called R Markdown

## R Markdown

R Markdown is a markup language (confusing, isn't it?) that combines `R` with a
plain text editor. There is a wealth of resources on how to use it and we will
be learning a lot about it but, for now, it's best to just see some basics and
play around with them.

You've already seen the boxes below in the first tutorial but now you have more
context to understand what they are all about. They show what R Markdown looks
like and how it gets "rendered" into rich text. Compare the two to see how the
markdown translates into formatting. As you will see, R Markdown makes it
pretty easy to create nicely formatted text using just a few symbols here and
there.

The top box is editable so feel free to type in anything you want.

```{r}
text <- "
# Level 1 heading

## Level 2 heading

### Level 3 heading, etc..

This is **bold**.

This is *italics*.

If you really want, you can also do ~~strikethrough~~.

One line break doesn't do anything.
These two sentences are still in the same paragraph.

For a new paragraph, we need two line breaks.

- this is
- an *unordered*
- list
    - it can have
    - multiple levels

1. this is
1. an *ordered*
1. list
    1. also with
    1. multiple levels

You can even write maths formulae! $\\sum_{i = 1}^{n}(x_i - \\bar{x})^2$
"
div <- htmltools::tag("div", list(class = "container"))
htmltools::tagAppendChild(div, text)
```

```{js}
document.body.onload = build;
function build (){
  let scriptTag = document.createElement('script');
  scriptTag.src = './js/main.js';
  document.body.appendChild(scriptTag);
}
```

```{css, include=TRUE}

pre code * {
 -ms-word-break: break-all;
     word-break: break-all;

     /* Non standard for WebKit */
     word-break: break-word;

-webkit-hyphens: auto;
   -moz-hyphens: auto;
        hyphens: auto;
}

.container {
  width: 100%;
}
textarea {
  overflow: hidden;
  resize: none;
  border: solid 1px rgb(50, 180, 70));
  min-height: 10em;
	font-family: monospace;
	width: 99.7%;
	padding: 1em;
	display: block;
	margin: auto;
}

textarea:focus {
	outline: solid 1px rgb(50, 180, 70);
}

.md-editor {
	padding: 3.5rem 0;
	position: relative;
}

.md-editor::before {
	font-family: 'Source Sans', sans-serif;
	color: rgb(50, 180, 70);
	content: 'Your markdown';
	position: absolute;
	margin-top: -18px;
	font-size: 2em;
	font-weight: bold;
	right: 0;
}

.rendered-content{
	position: relative;
	width: 100%;
  border: solid 1px rgb(67, 75, 117);
	background: #fff;
	padding: 1em;
	overflow-wrap: break-word;
}

.rendered-content::before {
	content: 'Rendered output';
	position: absolute;
	margin-top: -33px;
	font-size: 2em;
	font-weight: bold;
	right: 0;
}

```
 

Take a moment to play around with the editor to get used to the markdown. You
don't need to understand the symbolic language for the maths formula; it's just
to showcase what R Markdown can do. There is much more you can do in R
Markdown, including automatic references, including `R` code, plots, tables,
pictures, or even GIFs and videos. In fact, the vast majority of the teaching
materials for PAAS was written in R Markdown!

\ 

## R Markdown files and output documents

At this stage you are ready to learn how to write your own R Markdown-generated
documents.

As a first step towards this goal, watch the video below.

<iframe width="672" height="378"
src="https://www.youtube.com/embed/DDQxdxOAEXQ" frameborder="0"
allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope;
picture-in-picture" allowfullscreen></iframe>
([As always, there's also a transcript of the video if you need it](https://docs.google.com/document/d/1S5CVpVD2KvnCCHeMDwp32oNV4D5gywfEIE2vYXPY4o8/edit#))

\ 

You will be using R Markdown to write all your assignments for the next three
methods modules (PAAS, Analysing Data, Discovering Statistics) and we would
encourage you to adopt it for basically all writing, especially your
dissertation!

So, to get some practice, here are some tasks for you:

**1. Create a new RStudio project called `"week_03"` in your
`uni/psych_methods/paas` folder.**

**2. Create the minimal folder structure as discussed in the previous tutorial.**
You need the `"data"` and `"r_docs"` folders.

**3. Following the instructions in the video you just watched, create a new Rmd
file.**

**4. Change the author at the very top of the document to your name.**

**5. Change the title to "Lecture 3 notes".**

**6. Delete everything in the template file after the second set of `---`.**

**7. In the main body of the file, transcribe your Lecture 3 notes, making sure
they are nicely structured.** If you don't have notes, you are strongly advised
to re-watch the lecture and take notes. The same goes for every lecture.

**8. Save the file as `"lec_03_notes.Rmd"` in the `"r_docs"` folder.**

**9. Knit the file into a HTML document.**

**10. Admire how pretty the output is!**

\ 

That's all for this tutorial.

\ 
