---
title: 'Tables with `kableExtra` and plots with `ggplot2`'
output:
  rmarkdown::html_document
---

```{r setup,echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)
```

This document should be used in conjunction with the [Week 11 worksheet](https://paas.netlify.app/practicals/10_tables_and_plots/worksheet/).
Feel free to add text to write little notes to your future self.

You can knit this document whenever you want to see what the formatted HTML version looks like.

## Packages

```{r packages}
## load the required packages here
```


## Data

```{r data}
## edit code to read in the 2 data sets
# fill in the blanks, removing the ...
stroop_data <- readr::read_csv("...")
movie_data <- ...
```


## Tables


### Simple frequency table

```{r freq-tib}
peng_smry <- palmerpenguins::penguins %>%  # you can break line after each %>% if you want
  dplyr::group_by(.data = ..., ...) %>%
  dplyr::summarise(n = ...)
```


```{r freq-table}
peng_smry %>% ...
```


### Adding relative frequencies

```{r rel-freq-tib}
movie_smry <- movie_data %>%
```


### Table of summary statistics

```{r smry-table}
stroop_data %>%
```



## Plots


### Simple histogram

```{r base-layer}
# give the function the correct data and the right aesthetic (variable)
p1 <- ... %>%
  ggplot2::ggplot(data = ., aes(x = ...))
p1  # this prints out the plot
```

```{r geom-layer}
# add the histogram layer
p2 <- p1 + 
  
p2 # print out plot
```

```{r labs-layer}
# add the layer of axis labels, specifying both x and y labels
p3 <- p2 +
  labs(...)
p3 # print out plot
```


### Scatter plot

```{r scatterplot}
... %>% # provide data
  ggplot2::ggplot(aes(x = ..., y = ...)) + # map variables onto axes
  geom_point() +
  labs(...) # give axes proper labels
```


### Grouped scatterplot

```{r grouped-scatterplot}
... %>% #provide data
  ggplot(aes(...)) + # specify the 4 aesthetics needed
  ... + # add points with some transparency
  ... # add axis labels
```


### Violin-box plot combo

```{r viol-box}
# provide data
# pipe them into the base layer where you map your aesthetics
# add a violin plot layer with fill and transparency
# add a boxplot layer and change width
# add a layer of labels
# give the plot a caption
```


### Grouped violin plot (custom colours)

```{r grouped-viol-box}
# First, create the grouped plot and don't worry about custom colours and geom positions
# We'll take care of those next.
```


### Errorbar plot

```{r errorbar-data}
movie_smry_plot <- ...

```

```{r errorbar-plot}
movie_smry_plot %>%
  ggplot2::ggplot(...) +
  geom_errorbar(...(ymin = ..., ymax = ...), width = 0.04) +
  ...
```


### Publication-quality groupped errorbar chart {.death}

```{r showstopper}

```


