---
title: "Lecture 6: Samples, Populations, and Distributions"
subtitle: ""
author: "Dr Lincoln Colling"
date: "04 November 2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)

style_duo_accent(
  primary_color = "#003b49",
  secondary_color = "#1d4289",
  header_font_google = google_font("Cabin"),
  text_font_google = google_font("Noto Sans", "300", "300i"),
  code_font_google = google_font("Fira Mono"),
  colors = c(
    red = "#d3273e",
    purple = "#5d3754s",
    orange = "#dc582a",
    green = "#007a78",
    white = "#FFFFFF",
    blue = "#1d4289"
  )
)

xaringanthemer::style_extra_css(
  list(
    ".title-slide" = list(
      "background-image" = paste0(
        "url(https://upload.wikimedia.org/",
        "wikipedia/commons/thumb/3/34/",
        "University_of_Sussex_Logo.svg/",
        "480px-University_of_Sussex_Logo.svg.png)"
      ),
      "background-position" = "95% 95%",
      "background-size" = "180px",
      "border" = "10px solid #013035",
      "background-color" = "#FFFFFF"
    ),
    ".title-slide h1" = list(
      "padding-top" = "0px",
      "font-size" = "60px",
      "text-align" = "left",
      "padding-bottom" = "18px",
      "margin-bottom" = "18px",
      "margin-top" = "0px",
      "color" = "#003b49"
    ),
    ".title-slide h2" = list(
      "font-size" = "40px",
      "text-align" = "left",
      "padding-top" = "10px",
      "margin-top" = "0px",
      "color" = "#003b49"
    ),
    ".title-slide h3" = list(
      "font-size" = "30px",
      "color" = "#26272A",
      "text-align" = "left",
      "text-shadow" = "none",
      "padding" = "0px",
      "margin" = "0px",
      "line-height" = "1"
    ),
    ".footnote" = list(
      "font-size" = ".6em",
      "border-top" = "1px solid #007935"
    ),
    "h4" = list("color" = "#003b49")
  )
)
```

```{r, echo=FALSE}
library(metathis)
meta() %>%
  meta_description("Samples, populations, and distributions.") %>%
  meta_tag("week" = "06") %>%
  meta_tag("content_type" = "slides")
```


# Plan for today

Part I: **Samples and populations**
- The relationship between samples and populations

Part II: **Distributions**

- The **binomial** distribution

- The **normal** distribution

  - Describing the normal distribution, and processes that produce normal
    distributions

  - Processes that don’t produce normal distributions, and describing
    deviations from the normal distribution

Part III: **Distributions and samples**

---

## Samples and populations

A key use of **statistics** is  to make **inferences** (or **claims**) about
**populations** from *the information* we get from **samples**

**Example**

You're interested in the **.blue[average height]** of **.green[people in the
UK]**

*How can you go about collecting some data that will allow you to make claims
about the .blue[average height] of .green[people in the UK]?*

---

## Samples and populations

To you have (at least) two options:

**Option 1** 

- Measure the height of _.green[all the people in the UK]_ and then work out
  the .blue[average]

- But that's over 66 million people, so it'll take you a very long time and
  maybe some people don't want to be measured

**Option 2**

- Measure a *.red[subset]* of *.green[all the people in the UK]* and use the
  .red[average of this subset] to figure out plausible values for the
  .green[average height of people in the UK]

In this example, the .red[subset of people] is the .red[sample] and .green[all
the people in the UK] is the .green[population]

???

It's important to note that there isn't a **single** population. What counts as
the population will depend on the claim you're making. For example, let's say
I'm interested in testing the claim, "Do **people in East Sussex** show an
interference effect on the Stroop task?". Here the **population** would be
**people in East Sussex**. If, however, I want to make claims about **people in
general**, then the **population** might be **all living humans**. The
**sample** is always going to be a subset of the **population**. 

In the vast majority of psychology research tend to be rather **WEIRD**. That
is, the samples used in most published psychological science are drawn from
**W**estern, **E**ducated, **I**ndustrialized, **R**ich, and **D**emocratic
(WEIRD) societies [@weird]. The preponderance of WEIRD samples implies that
researchers either assume that there is little variation across different
population groups when it comes to the phenomena they're studying, or that
these samples are as representative of **humans in general** as any other
group.

---

### The relationship between samples and populations 

After we've taken a sample we'll want to use information from this
**.red[sample]** to figure out something about the **.green[population]**

_.blue[But what's the relationship between the population and the sample?]_

The **sample** should hopefully **resemble** the **population** in some way

- For example, the average of the sample should **resemble** the average of the
  population

- But we don't know the average of the population (if we did, then we wouldn't
  need the sample), so how would be _know_ whether our sample **resembles** the
  population?

We can do a _thought experiment_ to try and figure out some factors that will
influence whether the **sample** _resembles_ the **population**

---

### Relationship of sample to population

Let's think back to our question about _the average height of people in the UK_

#### Factor 1: Variation in the population

If all members of the population are **identical** then the height of one
person would be the same as the average height of two people, or 100 people, or
the entire population, because people only come in one height

_.blue[When there is no variation in the population]_ then the sample average
will be **identical** to the population average

---

### Relationship of sample to population

#### Factor 2: Size of the sample

If our sample is large enough so that it _includes all members of the
population_ the sample and the population are the same thing

*.blue[When the sample includes the entire population]* then the sample average
will be **identical** to the population average


<br />

These are extreme cases but they suggest that **population variation** and
**sample size** will influence the relationship between samples and populations

---

### Relationship of sample to population

So if we have a big sample and/or small population variation then will our
sample **resemble the population?**


**For a particular sample** there is no way of knowing whether it **resembles**
the population or not, because we don't know what the population looks like!


A better way to think about this is in terms of **repeated sampling**

- If we take lots of samples from the same population then will those samples
  **on average** be closer to the population?

- These two factors (sample size, and population variation) will influence
  whether the samples resemble the population **on average**

If our sample size is **big enough** then samples will **on average** resemble
the population...

...but what counts as **big enough** will depend on the **population
variation**


---

#### Repeated sampling from the same population

<iframe  width="700px" height="650px" style="border:none; margin-left:auto;
margin-right:auto;display:flex"
src="https://paas-embed.netlify.app/sampler/index.html?slides"
scrolling="no"></iframe>


---

## Distributions

Before we start talking about **distributions** let's think about what they are
and where they come from

_.blue[We'll do another thought experiment]_

- We'll take a coin, and we'll flip it.

- Two outcomes are possible
  1. The coin lands showing *heads*

  2. The coin lands showing *tails*

Of the two possible events

1. One produces 0 heads
2. One produces 1 head

Now let's add more coins. As we do, we'll count up the number of sequences that
produces 0 heads, 1 head, 2 heads, 3 heads etc


---

### The binomial distribution 

<iframe  width="700px" height="500px" style="border:none; margin-left:auto;
margin-right:auto;display:flex"
src="https://paas-embed.netlify.app/coins/index.html" scrolling="no"></iframe>

---

#### Plotting the frequency of outcomes


We'll treat the _number of heads in a sequence_ as our _outcome_

As we add more and more coins we can plot the **frequency** of each _possible
outcome_.

- This frequency plot starts to take on a **characteristic shape**

- This shape can be described mathematically using the **binomial
  distribution**

The **binomial distribution** describes the **frequency of outcomes** in our
coin flipping example<sup>1</sup> 


.footnote[<sup>1</sup>In our thought experiment we assume that every _possible_
sequence of Heads and Tails occurs, and that it occurs only once.]

---

#### Natural processes that produce binomial distribution

Balls falling through a _bean machine_ approximate a binomial distribution

<iframe  width="700px" height="660px" style="border:none; margin-left:auto;
margin-right:auto;display:flex"
src="https://paas-embed.netlify.app/bean/index.html?slides"
scrolling="yes"></iframe>

---

### The normal distribution

The shape seen in the **binomial distribution** is also seen in another
distribution called **the normal distribution**.

**Differences between the binomial distribution and the normal distribution:**

- The **binomial** distribution is **bounded** and the **normal distribution**
  is not

  - **The binomial distribution** ranges from 0 to n (where n is the number of
  coins you've flipped)

  - **The normal distribution** ranges from $-\infty$ to $+\infty$


- The **binomial** distribution is **discrete** and the **normal distribution**
  is **continuous**

  - You can only have 0 heads, 1 head, etc., and not 1.5 heads

  - *Normal distribution* represents all outcomes between $-\infty$ and
  $+\infty$


---

#### The normal distribution as a model

- The **normal distribution** is a mathematical abstraction (nothing in real
  life perfectly follows a **normal distribution**)

- But we can use it as a **model** of **real-life frequency distributions** 

<span class="pull-left" style="width:70%; margin-right:-130px"><iframe
width="800px" height="380px"
style="border:none;display:flex;align:left;margin-left:-200px"
src="https://paas-embed.netlify.app/simple_normal/index.html"
scrolling="no"></iframe></span>

The **normal distribution** can be described by two parameters:</span>

- The **&mu;** parameter controls where it is centred

- and the **&sigma;** parameter controls how wide it is.

---

#### Processes giving rise to normal distribution 

To see how a natural process can give rise to the **normal distribution** let's
consider a simple *dice game*

**The rules of the game**   

1. A group of **x** players roll a dice **n** times

2. A player's score is calculated by **adding** all the values of the **dice
   rolls** 

  - For example, if they rolled the dice three times (**n** = 3) and the dice
  showed **1**, **4**, and **4** then their score would be **9** (1 + 4 + 4 =
  9)

If you have enough dice rolls then the **players' scores will be normally
distributed** 

---

#### A dice game simulation 

<iframe  width="700px" height="400px" style="border:none; margin-left:auto;
margin-right:auto;display:flex"
src="https://paas-embed.netlify.app/dice/index.html?type=add&diceInput=30&playerInput=100"></iframe>


As you increase the number of **dice rolls** the **frequency distribution** of
_players scores_ will start to look like a normal distribution 

But you also need enough players to clearly see shape

---

#### Natural processes are analogous to the dice game

There are many natural processes that are **analogous** to the **dice game**

We can imagine other processes that work like the **dice game**

- For example, a developmental process might work similarly. 

  - At each point in time some value can be **added** on to the person’s
    current height just like players scores can increase by some amount on each
    dice roll.


**.green[The numbers you add aren't important... it's the _adding_ that's
important]**

---

#### Natural processes are analogous to the dice game

A person’s score can increase by either 1, 2, 3, 4, 5, or 6 after each roll,
and with a balanced dice an increase of 1 will be no more common than an
increase of 6, or 5 etc 

But even if the dice were unbalanced then a normal distribution would still
appear. 

The numbers that you add isn't the important thing... the **adding** is what's
important

If instead the numbers were **multiplied** then we wouldn't see a **normal
distribution**

---

#### A different dice game simulation 


<iframe  width="700px" height="400px" style="border:none; margin-left:auto;
margin-right:auto;display:flex"
src="https://paas-embed.netlify.app/dice/index.html?type=prod&diceInput=10&playerInput=30"></iframe>

We can change the rules of the dice game so that we **multiply** instead of
**add**

- This won't produce a **normal distribution**. 

- The distribution will be **skew** 


---

#### Describing deviations from the normal distribution

When distributions deviate from the normal distribution this can happen in two
ways 

- The distribution can be **asymmetrical**

- The size of the **tails** can change

.pull-left[ <iframe  width="700px" height="250px" style="border:none;
margin-left:auto; margin-right:auto;display:flex"
src="https://paas-embed.netlify.app/skew/index.html"></iframe>

- Asymmetry is quantified by **skew**]

.pull-right[ <iframe  width="700px" height="250px" style="border:none;
margin-left:auto; margin-right:auto;display:flex"
src="https://paas-embed.netlify.app/kurtosis/index.html"></iframe>


- The **thickness** of the **tails** is quantified by **kurtosis** (or **excess
  kurtosis** if given _relative_ to the **normal distribution**)]

---

## Distributions and samples

Now that we know a little about **distributions** we can return to **samples**

So far, with the dice game we've just been concerned with whether the scores of
the players **within** a game are normally distributed

But can we say anything about distribution of scores **across** games?

<br />

Let's return to the problem of **sampling** people and measuring their height

- After we've got our **sample** and we've taken our measurements let's **add**
  up all the measurements, and calculate the **total height** of our sample.

- Now let's take another sample and **add** all the measurements again

.blue[What can we say about how these **sums** will be distributed?]

Because we're taking **sums**, we know that the **total heights** will be
normally distributed!

---

## Distributions, samples, and averages

What if instead of taking a regular **sum** we can first *divide each value in
the sample by the sample size and then calculate the sum?

This turns the *sum* into an **average**, but because we're still dealing with
**sums**, the **averages** will be normally distributed too!

That is, we can expect that if we take lots of samples (select a group of
people and measure their height), and then we work out the **average** of each
sample, then these **sample averages** will be **normally distributed**

This will happen **irrespective** of how the **population is distributed**

- That is, even if **height** isn't **normally distributed** then the
  **average** height from lots of samples **will still be normally
  distributed**<sup>1</sup>



.footnote[<sup>1</sup>Remember, whether a dice shows 1, 2, 3, etc isn't
normally distributed. Each value is equally likely]

---

### Preview of the sampling distribution

The fact that **sample averages** are **normally distributed** underlies the
concept of the **sampling distribution**

The **sampling distribution** will underlie many of the statistical procedures
you'll learn about, and it'll be covered in more detail in the next lecture!




