---
title: "Introduction to computing"
subtitle: "Tutorial 01"
author: "Psychology as a Science"
output:
  rmarkdown::html_document:
    css: ["./css/paas_learnr.css", "./css/katex.min.css"] 
    includes: 
      js: "./js/setup.js"
      after_body: "./header.html"
    self_contained: false
---
 
```{r, include=FALSE}
library(metathis)
library(niceQuiz)
meta() %>%
  meta_description("Week 1 Tutorial") %>%
  meta_tag("content_type" = "tutorial",
           "week" = "01")
```

## PAAS tutorials


Welcome to the first Psychology as a Science tutorial! This is the Week 1
tutorial and this tutorial will get you prepared for the practical class in
Week 2.

This series of interactive tutorials will gently guide you through the topics
covered in the PAAS module. To get the best value out of this module, you
should complete each week's tutorial **ahead of the practical session**. These
tutorials are designed to take about an hour to complete. Some of them include
videos, some will guide you through the topic at hand with short exercises, and
some will offer you the opportunity to test your understanding with quiz
questions and tasks. It is in your best interest not to skip through these
tutorials and complete them all in due course.

We did our best to make the tutorials as fun as possible but that doesn't mean
you shouldn't take them seriously. The PAAS module covers a lot of topics you
might be completely unfamiliar with so making good use of all the resources
provided to you really is the best thing you can do to get the most out of the
module.

Let's get started!

Lincoln, Danielle, and Milan

 

## Exercises and quizzes

Before we start with this week's topic, let's have a look at some of the
exercise and quiz formats you will encounter in these tutorials.

### Text editor boxes

From week 3 on, we will start learning how to write documents in RStudio using
its text editor, called R Markdown. Unlike visual text editor that you might be
used to, such as Google Docs or Microsoft Word, R Markdown doesn't have any
buttons or icons you can press to make a part of the text **bold** or
*italicise* it. Instead, it uses plain text characters such as `*`, `_`, `#`,
`^`, and others to format the final document.


To help you get used to this way of writing documents, the tutorials will
sometimes let you practice R Markdown in text editor boxes such as the one
below. Anything you type in the box will get *rendered* underneath as it would
if you generated a document from a R Markdown file.

```{r, echo = FALSE}
text <- "
## Markdown example

This is **bold**.
This is *italics*.

- this is
- an *unordered*
- list

1. this is
1. an *ordered*
1. list

Even maths formulae! $\\sum_{i = 1}^{n}(x_i - \\bar{x})^2$
"
div <- htmltools::tag("div", list(class = "container"))
htmltools::tagAppendChild(div, text)
```

```{js, echo = FALSE}
document.body.onload = build;
function build (){
  let scriptTag = document.createElement('script');
  scriptTag.src = './js/main.js';
  document.body.appendChild(scriptTag);
}
```

\ 

Don't worry too much about what any of this means at this stage.
We will go through all of it step-by-step in a couple of weeks.


### Quiz questions

Many of the tutorials will also contain quiz questions that will allow you to check your understanding of the concepts covered in these tutorials.
They include single best answer questions, where you only select one options:

```{r echo = FALSE}
quiz(caption = "",
question("What is the name of the first module in the research methods and statistics block?",
         answer("Analysing Data"),
         answer("Qualitative and Quantitative methods"),
         answer("Psychology as a Science", correct = TRUE),
         answer("Discovering Statistics")
))
```

...and multiple choice questions, where your task is to check all correct options:

```{r echo=FALSE}
quiz(caption = "",
  question("Which of these are the core components of this module?",
    answer("Practicals", correct = TRUE),
    answer("Pub socials"),
    answer("Lectures", correct = TRUE),
    answer("Tutorials", correct = TRUE)))
```

No, they are not marked. &#128521;

## More features

The tutorials also have other features like code boxes, where you can
practice writing `R`. We'll come across these as the course progresses.

### R and RStudio

We will be using the RStudio software to write documents (*e.g.*, analysis
reports) and to process, analyse, and visualise data. RStudio is the most
popular software for programming in the `R` programming language. In order to
use it, you first need to install both `R` and RStudio.

**Your <a id="task1">first task</a>, install `R` and RStudio Desktop on your computer.**

To install `R`, you'll first need to download the correct version for your
system. The links are listed below:

- [`R` for Microsoft Windows](https://cloud.r-project.org/bin/windows/base/R-4.1.1-win.exe) 

- [`R` for an Intel Apple Mac](https://cloud.r-project.org/bin/macosx/base/R-4.1.1.pkg)

- [`R` for an M1 Apple Mac](https://cloud.r-project.org/bin/macosx/big-sur-arm64/base/R-4.1.1-arm64.pkg)

Once you have downloaded the install file, double-click it and proceed through
the install process (which will be similar to other software that you have
installed).

Once you've downloaded and installed `R` you will need to download and install
RStudio.

To download RStudio you can click the following links: 

- For the [MacOS version](https://download1.rstudio.org/desktop/macos/RStudio-1.4.1717.dmg)

- For the [Windows version](https://download1.rstudio.org/desktop/windows/RStudio-1.4.1717.exe)

If you need help installing either of these, here's a video guide on how to
install both on Windows:

<iframe title="YouTube video player"
src="https://www.youtube.com/embed/cLTXve3Rf5k" width="672" height="378"
allowfullscreen="allowfullscreen" allow="accelerometer; autoplay;
clipboard-write; encrypted-media; gyroscope; picture-in-picture"></iframe>

\ 

And here is one for MacOS users:

<iframe title="YouTube video player"
src="https://www.youtube.com/embed/TXxmVaZkYks" width="672" height="378"
allowfullscreen="allowfullscreen" allow="accelerometer; autoplay;
clipboard-write; encrypted-media; gyroscope; picture-in-picture"></iframe>

\

If you feel like you're just doing things and you don't understand what exactly
it is you're doing and why you're doing it, relax. We will start lifting the
fog of war pretty soon!

## Setting up your OneDrive folder (optional)

Each University of Sussex student is given a OneDrive account. This account
gives out quite a lot of cloud storage space (1TB).  If you want to access the
same files in the computer labs and on your home computer then you'll need to 
set up OneDrive on your home computer.

To set up OneDrive, log in using your University of Sussex credentials and sync
the folder with your computer.

**Your <a id="task2">second task</a> is to get access to your OneDrive folder
and sync it to your computer.** Only proceed with installing OneDrive if you
managed to install R and RStudio successfully. Installing OneDrive **before**
setting up R/RStudio can sometimes cause problems, particularly if you're using
Windows. If you had trouble installing R/RStudio or couldn't install it
correctly, then don't worry about installing OneDrive just yet. 

If you need help, the uni website has an [entire page on setting up and working
with OneDrive](http://www.sussex.ac.uk/its/onedrive).

## Computers

**Your <a id="task3">third task</a> is to watch/listen to this video tutorial.**

It covers some basic principles of computer programming. We recommend you watch
it in the HD resolution in a full screen mode.

<div class="accessible"> The video below contains a flowchart illustrating an
important concept on the example of making a cup of tea. For screen reader
users, below is the transcript of this flowchart.

1. Go to cupboard.
1. Open cupboard.
1. Search for mug.
1. If there is mug go to 10. 
1. If no mug, search for glass.
1. If no glass, go to 52.
1. Get glass.
1. Call glass “receptacle”.
1. Go to 12.
1. Get mug.
1. Call mug “receptacle”.
1. Go to tea cupboard.
1. Open tea cupboard.
1. Search for tea bag.
1. If no tea, go to 52.
1. Get tea bag.
1. Put tea bag in receptacle.
1. Go to kettle.
1. If more than 500ml water in kettle, go to 30.
1. Take kettle.
1. Go to kitchen sink.
1. Open kettle lid.
1. Place kettle under tap.
1. Open tap.
1. If more than 500ml water in kettle, go to 27.
1. Go to 25.
1. Turn off tap.
1. Close kettle lid.
1. Go to kettle base.
1. If kettle is on base go to 32.
1. Place kettle on base.
1. If kettle plugged in the mains, go to 34.
1. Plug kettle in the mains.
1. Switch on kettle.
1. If kettle is off, go to 37.
1. Go to 35.
1. Take kettle.
1. Go to receptacle.
1. Pour water into receptacle.
1. If more than 250ml of water in receptacle, go to 42.
1. Go to 40.
1. Stop pouring water.
1. Start timer.
1. Go to kettle base.
1. Place kettle on kettle base.
1. Go to receptacle.
1. If timer shows more than 4 minutes, go to 49.
1. Go to 47.
1. Remove tea bag from receptacle.
1. Go to bin.
1. Discard tea bag.
1. Exit.
</div>

<iframe width="672" height="378"
src="https://www.youtube.com/embed/ockaUxLtc_c" frameborder="0"
allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope;
picture-in-picture" allowfullscreen></iframe>

If you need it, there is also a [transcript of the video
available](https://docs.google.com/document/d/1yKTSaDoi-jacVGvMLBGKR52KCOEXkX8sG1oyD8mwR4s/edit?usp=sharing).


**Your <a id="task4">fourth and final task</a> is to set up your RStudio according to the
instructions in the video.**

Once you've finished installing RStudio, make sure you open it up and run the
following command at the **console**. This command will install everything you
need for this course.

```{r, class.source="copy", eval=FALSE}
source("https://paas.netlify.app/install.r")
```

If you've had any problems installing anything from this tutorial then feel
free to come to one of the drop-in sessions or bring your laptop to an
in-person practical.

You should now have performed all four tasks:

1. [Installed `R` and RStudio Desktop on your computer](#task1)

2. [Set up OneDrive on your computer](#task2) (optional)

3. [Wacthed the tutorial video](#task3)

4. [Set up RStudio according to the instructions in the video](#task4)

That is all for this week.

\
